---
title: "2.07_Volcano-Plots"
author: "Andrew Davis"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_document:
    toc: TRUE
---
# Objectives:
In this script the user will generate volcano plots for each comparison results generated in [2.03-DESeq2.Rmd](../R_scripts/2.03-DESeq2.Rmd)

### Install/Load dependencies
```{r, echo=TRUE, message=FALSE, Loading_Dependencies_205}
# Load dependencies:
library('ggplot2')
library('EnhancedVolcano')
library(magrittr)
```

### Plot volcanoes:
The following function takes 2 input arguments:
  1. DEGs: the results data frame output from the "annotate" function in 2.03-DESeq2.Rmd
  2. filenames: a character string describing the particular comparison being made in the DEGs df
  
Note that the x and y axes limits may need to be modified depending on the DE results for your particular experiment
```{r volcano_function}
##identify max p-value and log2FC in comparisons:
adjP_vals<-lapply(anno_results,function(i) i[,7]) %>% do.call(cbind,.) %>% data.frame
logFC<-lapply(anno_results,function(i) i[,4]) %>% do.call(cbind,.) %>% data.frame


min_adjP<-min(adjP_vals[,1],na.rm=TRUE)
min_adjP<--log(min_adjP,base=10)
max_logFC<-max(logFC[,1],na.rm=TRUE)
min_logFC<-min(logFC[,1],na.rm=TRUE)

plot_volcano <- function(DEGs,filenames){
pdf(sprintf("../Results/%s_volcano_padj-0.1_log2FC-1.pdf",filenames), height=8, width=8)
  volcano <- EnhancedVolcano(DEGs,
        lab = DEGs$gene_name,
        x = 'logFC',
        y = 'adj.P.Val',
        xlab = bquote(~Log[2]~ "fold change"),
        ylab = bquote(~-Log[10]~adjusted~italic(P)),
        pCutoff = 0.1,
        FCcutoff = 1,
        xlim=c((min_logFC-2),(max_logFC+2)),
        ylim=c(-0.1,(min_adjP+2)),
        colAlpha = 1,
        legendPosition = "bottom",
        legendLabSize = 14,
        legendIconSize = 4.0,
        border = "full",
        borderWidth = 1.5,
        borderColour = "black",
        gridlines.major = FALSE,
        gridlines.minor = FALSE,
        title = sprintf("%s",filenames))
  print(volcano)
dev.off()
  print(filenames)
  print(volcano)
}
```

### plot volcanoes
```{r Plot-volcanoes}
# Annotate and export the volcano plots:
for(i in 1:length(comparisons)){
plot_volcano(anno_results[[i]],paste0(comparisons[[i]][2],"_vs_",comparisons[[i]][3]))
}
```

### Next step: 
Proceed to run the .Rmd script: [3.03_GSEA.Rmd](3.03_GSEA.Rmd) summarized in [3.03_GSEA.html](../Results/3.03_GSEA.html)

### Session information
```{r session_info_207}
sessionInfo()
```

This document was processed on: `r Sys.Date()`.