---
title: "3.03_GSEA"
author: "Andrew Davis"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_document:
    toc: TRUE
---

# Objectives:
In this script the user will run fGSEA against GO, Reactome, and Wikipathways on the DESeq2 results for each comparison generated in [2.03_DESeq2.Rmd](../R_scripts/2.03_DESeq2.Rmd)

# Install/Load dependencies
```{r include=FALSE}
# Load dependencies:
library('readxl')
library('dplyr')
library('fgsea')
library('org.Hs.eg.db')
library('GO.db')
library("reactome.db")
library('ggplot2')
library('GSEABase')
library('data.table')
library('biomartr')
library('annotate')
library('msigdbr')
```

## Set random seed
Setting a random seed helps keep the scripts reproducible
```{r set_random}
set.seed(1234)
```

## Download gene sets:
```{r, Download_pathways_msigdbr}
# Retrieve human genes for all gene sets in the database.
m_df_BP = msigdbr(species = "Homo sapiens",subcategory=c("BP"))
m_df_MF = msigdbr(species = "Homo sapiens",subcategory=c("MF"))
m_df_CPREACTOME = msigdbr(species = "Homo sapiens",subcategory=c("CP:REACTOME"))
m_df_wikipathways = msigdbr(species = "Homo sapiens",subcategory=("WIKIPATHWAYS"))
m_list_BP = m_df_BP %>% split(x = .$gene_symbol, f = .$gs_name)
m_list_MF = m_df_MF %>% split(x = .$gene_symbol, f = .$gs_name)
m_list_CPREACTOME = m_df_CPREACTOME %>% split(x = .$gene_symbol, f = .$gs_name)
m_list_wikipathways = m_df_wikipathways %>% split(x = .$gene_symbol, f = .$gs_name)
m_list<-c(m_list_BP,m_list_MF,m_list_CPREACTOME,m_list_wikipathways)
```

## Functions:
```{r, Functions_5}
# Rank the genes by adj p-value and log2-FC
rank_DEG <- function(DEGs){
  DEGs_ranked <- DEGs
  DEGs_ranked$fcsign <- sign(DEGs_ranked$logFC)
DEGs_ranked[6] <- lapply(DEGs_ranked[6], function(x) replace(x, x == 0, min(x[x>0], na.rm = TRUE)))
  DEGs_ranked$logP<--log(DEGs_ranked$pvalue)
  DEGs_ranked$metric= DEGs_ranked$logP/DEGs_ranked$fcsign
  # Prepare ranked list for running GSEA
  # Subset for gene_name and rank metric
  y<-DEGs_ranked[,c("gene_name", "metric")]
  y<-as.data.frame(y)
  y_uni = y[!duplicated(y$gene_name),]
  # convert y to a named vector, which is required format for fgsea
  y_uni<-setNames(y_uni$metric,y_uni$gene_name)
  # sort the list
  y_uni_sort<-sort(y_uni)
  return(y_uni_sort)
}
# Perform GSEA of the full preranked gene list against Reactome and GO pathways and plot the results:
GSEA_ALL <- function(ranked_DGEs,filename){
  # Run GSEA for Reactome pathways
  fgseaRes <- fgseaMultilevel(pathways=m_list, stats=ranked_DGEs, minSize=15,maxSize=500)
  # Export the ranked gene sets:
  fgseaRes_ex <- fgseaRes
  fgseaRes_ex$leadingEdge <- vapply(fgseaRes_ex$leadingEdge, paste, collapse = ", ", character(1L))
  write.csv((fgseaRes_ex[order(fgseaRes_ex$padj,-abs(fgseaRes_ex$NES)), ]),sprintf("../Results/GSEA_%s.csv",filename))
}
```


## Run GSEA on all comparisons:
```{r}
for(i in 1:length(comparisons)){
  # Rank the genes by p-value and log2-FC
DEG_sort <- rank_DEG(anno_results[[i]])
# Perform GSEA of the full preranked gene list against Reactome and GO pathways and plot the results
GSEA_ALL(DEG_sort,paste0(comparisons[[i]][2],"_vs_",comparisons[[i]][3]))
}
```

### Next step: 
Proceed to run the .Rmd script: [4.03_Export-to-IPA.Rmd](4.03_Export-to-IPA.Rmd) summarized in [4.03_Export-to-IPA.html](../Results/4.03_Export-to-IPA.html)

### Session information
```{r session_info_303}
sessionInfo()
```

This document was processed on: `r Sys.Date()`.