---
title: "4.03_Export-to-IPA"
author: "Andrew Davis"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_document:
    toc: TRUE
---

# Objectives:
In this script the user will collate all of the DESeq2 results for each comparison generated in [2.03_DESeq2.Rmd](../R_scripts/2.03_DESeq2.Rmd), which is useful for uploading into IPA.

### Install/Load dependencies
```{r, echo=TRUE, message=FALSE, Loading_Dependencies_203}
# Load dependencies:
library(rlist)
```

## Concatenate all of the results into a single table and export (for IPA, etc):
```{r}
# Create the list of DEG data frames:
new_names<-list()
for(i in 1:length(comparisons)){
new_names<-append(new_names,list(paste0(comparisons[[i]][2],"_vs_",comparisons[[i]][3])))
}
DESeq2_list_suffix <- lapply(1:length(anno_results), function(i) {
  suffix <- anno_results[[i]] # Double brackets for a list
  names(suffix) <- paste0(names(suffix), sep=".",paste0(comparisons[[i]][2],"_vs_",comparisons[[i]][3])) 
  suffix$gene_id<-row.names(suffix)
  suffix 
})
# For Full (Outer) Join
res_all <- Reduce(
  function(x, y, ...) merge(x, y, all = TRUE,by="gene_id"),DESeq2_list_suffix
)
# Fix column names (change the last number (width of the dataframe))
res_all<-res_all[,c(1,3,4:ncol(res_all))]
colnames(res_all)[c(2,3)]<-c("gene_name","base_mean")
# for the sake of IPA, I converted all of the NA adjP values to 1:
res_all[is.na(res_all)] <- 1
write.csv(res_all,"../IPA/Merged_Treatment_Comparisons_DESeq2_Results.csv")
```

## Generate DEG summary table:
```{r DEG_tables}
DEG_columns<-c("Comparison","Total DEGs (adjP<0.1)","Upregulated genes (logFC>0, adjP<0.1)","Downregulated genes (logFC<0, adjP<0.1)")
# for loop
DEG_sig<-as.list(0)
DEG_row<-as.list(0)
for (i in 1:length(anno_results)){
DEG_sig[[i]]<-subset(anno_results[[i]],anno_results[[i]]$adj.P.Val<0.1)
DEG_row[[i]]<-c(as.character(new_names[i]),length(DEG_sig[[i]]$logFC),length((subset(DEG_sig[[i]],DEG_sig[[i]]$logFC>0)$logFC)),length(subset(DEG_sig[[i]],DEG_sig[[i]]$logFC<0)$logFC))
}
DEG_table<-list.rbind(DEG_row)
colnames(DEG_table)<-DEG_columns
write.csv(DEG_table,"../Results/DEG_Summary_Table.csv",row.names = F)
```

### Next step: 
Proceed to run the .Rmd script: [5.03_Gene-Heatmaps.Rmd](5.03_Gene-Heatmaps.Rmd) summarized in [5.03_Gene-Heatmaps.html](../Results/5.03_Gene-Heatmaps.html)

### Session information
```{r session_info_403}
sessionInfo()
```

This document was processed on: `r Sys.Date()`.